A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-06-23 tex/BiSig.tex
[ tex ] Most of S2
[ tex ] S1: related work on language-formalisation frameworks

%\usepackage[inline]{enumitem} % for environment enumerate*
%\setlist[enumerate]{mode=unboxed}

@@

\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40,caption={}]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40,caption={}]{{#1}}}

@@

  In response to this difficulty, we develop a \emph{simple} yet \emph{general} and \emph{constructive} theory of bidirectional type synthesis where \emph{bidirectional type synthesis} refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

@@

\input{sec1-intro}
\input{sec2-outline}
\input{sec3-defs.tex}
\input{sec4-bidirectionalisation.tex}
\input{sec5-synthesis.tex}
\input{sec6-formalisation.tex}
\input{sec7-future.tex}

>>>>

\usepackage[inline]{enumitem} % for environment enumerate*
%\setlist[enumerate]{mode=unboxed}

@@

\newcommand{\LT}[1]{\todo[author=LT,inline,color=orange!40]{{#1}}}
\newcommand{\Josh}[1]{\todo[author=Josh,inline,color=yellow!40]{{#1}}}

@@

  In response to this difficulty, we develop a simple yet general and constructive theory of bidirectional type synthesis where bidirectional type synthesis refers to the mutual use of type checking and inference based on bidirectional typing rules to determine the type of a program with type annotations.
  First, our theory is simple yet general as it deals with any simple type theory and its bidirectional type system that can be specified by a signature, analogous to the role of grammar in parsing.
  Second, our theory is constructive because it is not only formalised in a proof assistant based on constructive type theory but also formulated positively in order to compute.
  Specifically, the `proof' of logical decidability of type synthesis is instantiated to a synthesiser for the simple type theory specified by a valid signature, analogous to a parser generator.
  We introduce the notion of bidirectionalisation to refine annotatability in bidirectional typing.
  Type synthesis and bidirectionalisation result in the trichotomy of raw terms, which models a practical type synthesiser that may throw an exception and locate missing annotations.
  By formalizing our general theory of bidirectional type synthesis constructively, we provide a correct-by-construction `type-checker generator' for simple type theories, taking the first step towards a more general theory of bidirectional type synthesis.
\end{abstract}

@@

\input{sec1-intro}
\input{sec2-outline}
\input{sec3-defs.tex}
\input{sec4-pre-synthesis.tex}
\input{sec5-synthesis.tex}
\input{sec6-formalisation.tex}
\input{sec7-future.tex}

### 2023-06-23 tex/sec1-intro.tex
[ tex ] S1: related work on language-formalisation frameworks

%  Bidirectional typing had only been a set of principles and characteristics to design typing rules so that a type synthesis algorithm can be `read off' easily
  %, subsuming soundness, completeness, and decidability of type synthesis algorithm.
%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.

  %In short, this paper is a small but first step towards a general theory of type checking and inference.
  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
  %Our work addresses the problem with existing methodologies, 
Points to elaborate:
\begin{enumerate}
  \item Type inference and checking as a transition from (parsed) abstract syntax trees to well-typed terms.
  \item Bidirectional typing is by now the only solution that scales well.
  \item Not yet a general theory but by now a design principle.
    No `type checker generator' as each type checker is implemented on a case-by-case basis, but an algorithm can usually be read off from syntax-directed typing rules.
    A hint to a general pattern in syntax-directedness.
  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
    Lightly or rarely discussed.
  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
    Soundness of completeness of a type checking algorithm in a less straightforward way.
  \item A review of what has been done towards this line of research.
  \item Finally, our responses (contributions of this paper) and some example.
\end{enumerate}

\begin{enumerate}
  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
  \item Argue that syntax-directed form is essential and practical.
  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
\end{enumerate}

\subsection{Related work}

\citep{Xie2018}
\paragraph{Language formalization and its frameworks}
\cite{Wadler2022}
PoplMark Challenge~\citep{Aydemir2005}

\cite{Cimini2020,Cimini2022}

\citep{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022}
\cite{Allais2021}

@@

\paragraph{Bidirectional typing}

\cite{Pierce2000}
\cite{Peyton-Jones2007}
\cite{Dunfield2021}

@@

Our restrictions:
\begin{enumerate}
  \item Assumptions: Why only simple types? Various ways to deal with polymorphism; consistent on simple types.
  \item Assumption: Syntax-directed typing rules
\end{enumerate}

>>>>

Type checking and inference serve as the transition from parsed abstract syntax trees, referred to as \emph{raw terms}, to well-typed programs, known as \emph{well-typed terms}.
Type inference algorithms were developed for determining the type of any unannotated raw term, but it was discovered that full parametric polymorphism makes the type inference problem undecidable~\citep{Wells1999} as well as dependent types~\citep{Dowek1993}.
Recognising the limitations of type inference, the concept of bidirectional type synthesis based on bidirectional typing emerged as an alternative approach, offering decidable algorithms for determining the types of \emph{suitably annotated} programs in programming languages with polymorphic types, dependent types, gradual types, and more.

The core idea of bidirectional type synthesis is simple.
First we start with bidirectional typing which extends typing judgements with two modes that govern the flow of types, hence its name.
That is, typing judgements are classified into two forms:
  \[
    (\text{synthesising})\quad \Gamma \vdash \isTerm{t} \syn A
    \quad\text{and}\quad
    (\text{checking})\quad \Gamma \vdash \isTerm{t} \chk A
  \]
where the type of a term can be synthesised as the output from the information in its context and the term itself as inputs; it can be checked against a type additionally as an input, depending on \emph{the mode of a raw term}.
Then, an algorithm of bidirectional type synthesis can be `read off' based on how type information flows, if the output can be determined by inputs without guessing by design.
  For example, the type of a raw term $\isTerm{t}$ may be synthesised by looking up the variable in its context if $\isTerm{t}$ is a variable or by using a type annotation $A$ provided in the term if $\isTerm{t}$ is of the form $t \annotate A$ after checking $\isTerm{t}$ against the given type~$A$ first, ensuring the validity of the provided annotation.

Type annotations may be seen burdensome at first, but in fact they are useful for programmers to understand the purpose of a program.
This claim was put forth by~\citet{Pierce2000} and supported by software engineering practices, since type annotations are often included as part of top-level definitions and of the documentation even for languages such as variants of Hindley--Damas--Milner type system where (complete) type inference is possible.
As a result, the focus of type-checking technology has shifted from type inference to bidirectional type synthesis, considering both theoretical limitations and practical considerations.

  A recent survey paper by~\citet{Dunfield2021} summarised the design principles of bidirectional typing and its coverage of applications so far.
  Bidirectional type synthesis does not require unification---the traditional technique used in type inference.
Also argued by~\citet{Pierce2000}, missing annotations in bidirectional type synthesis can be recovered using the local information only, so introducing long-distance unification constraints to bidirectional type synthesis actually breaks its premise of being local.
 We may rightly claim that bidirectional type synthesis is the only approach that works in theory and in practice for a variety of languages.

Unlike parsing, however, which has theory and practice that work for a wide class of languages, bidirectional typing has only been developed on a case-by-case basis, lacking a general theory.
Although it is rather easy to `read off' an algorithm of bidirectional type synthesis, a theory that provides explicit specifications with rigorously proven properties is still absent.
By now, we can only offer informal design principles and considerations for individual systems~\citep{Dunfield2021}.
Likewise, there is no off-the-shelf `type-checker generator', and each type checker based on bidirectional typing has to be implemented independently.

To address the aforementioned issue, this paper aims to develop a theory of bidirectional type synthesis and a type-checker generator.
While competing ways exist for designing typing rules, such as those related to polymorphic types by~\citet{Pierce2000,Peyton-Jones2007,Dunfield2013,Xie2018} to name a few, the derivation of an algorithm is relatively standard once a set of bidirectional typing rules are in place.
Therefore, the goal of our theory is not to formulate bidirectional typing for different language features, but rather to define the `grammar' of bidirectional typing rules, referred to as the \emph{bidirectional binding signature} in this paper, along with its essential criteria including \emph{mode-correctness}, \emph{completeness}, and \emph{annotatability} in bidirectional typing that are adequate for deriving a type synthesis algorithm as outlined informally \emph{op.\ cit}.
To illustrate the core idea, we confine our discussion to simple type theories and their bidirectional type systems that have syntax-directed typing rules for simplicity, while more advanced type theories are left for future work.

\subsection{Related work}
As we will mechanise our theory not only for all simple type theories that can be specified by some signature but also with perspectives and techniques different from existing approaches in bidirectional typing, it is timely to take a review of the related work upon which this work is built.

\paragraph{Language formalization and its frameworks}
Formalising the metatheory of a programming language was initiated by the \PoplMark challenge~\citep{Aydemir2005} and this objective has been manifested in a textbook by~\citet{Wadler2022} where programming language concepts are all discussed formally in the proof assistant~\Agda, including bidirectional type synthesis for~\PCF.

It was soon observed that rudimentary operations and meta-properties for variable binding and substitution constitute the bulk of language formalisation, so it is rather unwise to tackle language formalisation on a case-by-case basis.
Hence, there are language-formalisation frameworks by~\citet{Ahrens2018,Fiore2022,Gheri2020,Ahrens2022,Allais2021} to establish substitution, term traversal, and their meta-properties once and for all supported abstract syntax with variable binding.
One of the core ideas of these frameworks is the concept of \emph{descriptions} (coined by~\citet{Chapman2010} for datatype-generic programming) or \emph{binding signatures} (coined by~\citet{Aczel1978} in line with the term in universal algebra) for specifying the set of typing rules of a language.
It is noteworthy that these frameworks only work for simple types at most and none of them can express polymorphic types nor dependent types.

\begin{remark}\label{re:type-synthesis-as-decidability-proof}
It is interesting to note that the algorithm of bidirectional type synthesis by~\citeauthor{Wadler2022} is formulated very differently from the literature:
\begin{enumerate*}
  \item the former algorithm is presented as a (logical) decidability proof of whether a bidirectionally decorated raw term can be bidirectionally typed or equivalently a dependently typed program that takes a raw term $\isTerm{t}$ and returns a typing derivation for $\isTerm{t}$ if there is one or the evidence of a contradiction otherwise;

  \item an algorithm in the latter form is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, which is read that type annotations can be added to $\isTerm{t}$ in the external language to yield $\isTerm{t}'$ of type $A$ in the internal language, and accompanied with the \emph{soundness} and \emph{completeness} statements.
\end{enumerate*}
\end{remark}

@@

@@

Moreover, given this opportunity of re-examining elements of bidirectional typing, we begin our theory from a general idea of type synthesis independent of bidirectional type synthesis and spot subtleties that are rarely, if any, discussed in the literature on bidirectional typing.

Our approach involves constructing the theory within Martin-L\"of type theory and formalising it using the \Agda proof assistant.
This formalisation provides a generic type synthesis algorithm that is applicable to all calculi employing bidirectional typing for simple types, hence essentially a verified type-checker generator.
By rigorously defining its core principles, we aim to lay the foundation for further advancements in bidirectional typing.

%  Our work addresses the problem with existing methodologies, setting the stage for further advances towards a more general theory of type synthesis.

  %By formalizing our theory of bidirectional type synthesise in \Agda, we provide a framework capable of handling bidirectional type systems of simple types uniformly.
  %Our work addresses the problem with existing methodologies,

Points to elaborate:
\begin{enumerate}
  \item The existing work of bidirectional typing does not take account of practical aspects such as missing annotation.
    Lightly or rarely discussed.
  \item A meta-problem: existing work is based on classical logic (or left unspecified at least).
    Soundness of completeness of a type checking algorithm in a less straightforward way.
\end{enumerate}
Assumption to justify:
\begin{enumerate}
  \item Use application in spin form to justify our design choice: type expression of arbitrary rank and the number of constructs in a calculus.
  \item Argue that syntax-directed form is essential and practical.
  \item Pfenning's recipe is only of design guide not a technical requirement (there are many exceptions).
\end{enumerate}
